---
jupyter:
  jupytext:
    formats: ipynb,Rmd
    text_representation:
      extension: .Rmd
      format_name: rmarkdown
      format_version: '1.1'
      jupytext_version: 1.2.4
  kernelspec:
    display_name: R [conda env:rkernel] *
    language: R
    name: conda-env-rkernel-r
---

# Honours Project - Expression Quantification
To identify expression levels, TPM (transcripts per million) can be used. (1)\
A slightly altered calculation has been proposed that may be better suited to compare expression levels between samples. (2)\
Both calculations require transcript length. Transcript length was retrieved from a gtf file using refGenome.
1. Li, B., Ruotti, V., Stewart, R. M., Thomson, J. A., & Dewey, C. N. (2009). RNA-Seq gene expression estimation with read mapping uncertainty.\
Bioinformatics, 26(4), 493–500. https://doi.org/10.1093/bioinformatics/btp692
2. Wagner, G. P., Kin, K., & Lynch, V. J. (2012). Measurement of mRNA abundance using RNA-seq data: RPKM measure is inconsistent among samples.\
Theory in Biosciences, 131(4), 281–285. https://doi.org/10.1007/s12064-012-0162-3

```{r}
countdf <- readRDS("data/devspine_counts_excluded.rds")
```

```{r}
require(refGenome)
require(stringr)
genome <- ensemblGenome()
read.gtf(genome, filename="data/Mus_musculus.GRCm38.87.gtf") #opens the gtf file
ann <- getGenePositions(genome,"gene_id") #retrieves gene annotations from the gtf file
ann["gene_id"] <- str_to_lower(ann[["gene_id"]])
ann <- ann[,colSums(is.na(ann))<nrow(ann)] #drops empty columns
geneann <- data.frame(ensembl_id=rownames(countdf)) #create new dataframe that will have genes in the right order
geneann <- merge(geneann, ann, by.x = "ensembl_id", by.y = "gene_id", all.x=TRUE, all.y=FALSE, sort=FALSE) #merge gene annotations dataframe with new one to order it
geneann["length"] <- geneann[["end"]] - geneann[["start"]] + 1 #calculate the length of each gene
rm(ann) #delete obsolete dataframe
saveRDS(geneann, file="data/devspine_genes.rds")
```

```{r}
calcTPM <- function(counts, lengths) { #This function calculates TPM as described by Li et al.
    RPB = counts/lengths
    TPM = RPB/sum(RPB)*10^6
    return(TPM)
}

altcalcTPM <- function(counts, lengths, readlengths) { #This function implements the alternative TPM calculation as described by Wagner et al., but, for now, the data for readlengths is not available.
    CLPL = (readlengths*counts)/lengths
    TPM = CLPL/sum(CLPL)*10^6
    return(TPM)
}
```

```{r}
TPM <- apply(countdf, 2, calcTPM, geneann[["length"]])
```

```{r}
saveRDS(TPM, file="data/TPM.rds")
```

The TPM values were saved.
